---
title: "3D geometric processing for dimensional control with dimControl"
author: "Nataly Romaris, Olamar Benavente, Ruben Fernandez-Casal, Salvador Naya, Javier Tarrio-Saavedra"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3D geometric processing for dimensional control with dimControl}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "##",
  # fig.dim = c(6, 5),
  fig.align = "center"
)
library(rgl)
setupKnitr()
```

## Introduction

`dimControl` provides tools for the geometric processing of three-dimensional point clouds and triangular meshes, with particular application to dimensional inspection.

This vignette illustrates a reproducible workflow for the processing of 3D geometric data using a reference CAD model included in the package as an example. A synthetic point cloud is first generated from the reference geometry to reproduce a simplified 3D scanning scenario. The object points are then separated from the supporting floor and used to reconstruct a triangular surface.

The reconstructed mesh is subsequently processed to identify and classify relevant geometric components according to their characteristics. Finally, the boundary of the object is extracted, cleaned, ordered, and segmented.

The workflow therefore covers:

- Simulation and preprocessing of a 3D point cloud.
- Reconstruction of a surface from a point cloud.
- Identification and classification of geometric components.
- Extraction, cleaning, and segmentation of boundaries.

```{r setup}
library(dimControl)
```

The [`rgl`](https://dmurdoch.github.io/rgl/) package is used for interactive 3D
visualizations, while [`Rvcg`](https://cran.r-project.org/package=Rvcg) package provides tools for processing triangular meshes, including the computation of face normals and barycentres.

Additionally, the [`legendplot`](https://rubenfcasal.github.io/legendplot/)
package provides complementary tools for 3D visualization, including functions
for managing the 3D display, adjusting viewpoints, and adding legends to graphical representations.

```{r message=FALSE}
library(legendplot)
```

The `legendplot::new3d()` function can be used to initialize or reset the 3D display.

## Reference geometry

A dimensional control workflow may include a reference geometry, such as a theoretical CAD model, when one is available. However, a reference CAD model is not required to use the geometric processing and measurement tools provided by `dimControl`.

When available, a CAD model can provide the theoretical reference geometry for dimensional measurements and can subsequently be compared with the corresponding geometry reconstructed from the inspected object.

In this vignette, the CAD model included in `dimControl` is used as an example to illustrate the workflow.

```{r show-cad, webgl=TRUE}
# Represent the CAD model
# Use `open3d()` or `new3d()` to open a new device
shade3d(cad, col = "lightgray")
```

```{r include=FALSE}
clear3d()
```

## Simulating a 3D point cloud from the CAD model

In a real dimensional inspection workflow, the geometry of the inspected object can be acquired using a 3D scanning system, producing a point cloud that represents the measured surface. Depending on the acquisition conditions, the resulting point cloud may also contain points belonging to the surrounding environment.

For this reproducible example, a synthetic point cloud is generated from the reference geometry using `simulateCloud()`. The simulated data contain points representing both the object and the supporting floor, reproducing a simplified 3D scanning scenario.

Point clouds can be very large. We have worked with point clouds containing more than 100 million points. However, a smaller point cloud is used in this example to reduce memory requirements and computational time, particularly during package checks.

```{r simulate-cloud}
set.seed(123)
cloud <- simulateCloud(mesh = cad, n = 1e5, rgen = rtnorm, mean = 0, sd = 0.2,
                       a = -0.5, b = 0.5, floor = list(n = 5e3, gap = 2, margin = 20))
head(cloud)
```

## Point cloud preprocessing

The simulated point cloud contains both object and floor points. Before surface reconstruction, the point cloud is preprocessed to isolate the points belonging to the object.

### Separating object and floor points

When the object and unwanted elements of the surrounding environment occupy different ranges along a coordinate axis, they can be separated by analyzing the distribution of that coordinate. The `findModes()` function identifies modes and valleys in this distribution, which can be used to determine a threshold for separating the components.

In this example, the Z coordinate is used to separate the object from the supporting floor.

```{r find-modes}
modesZ <- findModes(x = cloud[, 3], bw = 0.01, q1 = 0.97)
modesZ
```

The selected threshold is then used to classify each point as belonging to either the object or the floor.

```{r segment-cloud}
component <- factor(cloud[, 3] > modesZ[1, "right"], levels = c(FALSE, TRUE),
                    labels = c("floor", "object"))
table(component)
```

The resulting factor identifies the component associated with each point in `cloud`. The segmented point cloud is represented below, with object points shown in green and floor points in red.

```{r show-segmentation, rgl=TRUE}
# Represent the segmented point cloud with a categorical legend
fplot3d(levels(component))
points3d(cloud, lit = TRUE, col = fcolor(component), alpha = 0.1)
```

```{r include=FALSE}
clear3d()
```

The object points are retained for the subsequent surface reconstruction.

## Reconstructing the object surface

After separating the object from the supporting floor, the object point cloud is used to reconstruct a continuous triangular surface. The `createMesh()` function performs linear binning of the point cloud and generates a isosurface using the Marching Cubes algorithm.

In this example, the reference CAD model is provided through the `cad` argument. Its dimensions are used as a reference for the spatial coverage of the reconstructed surface. In addition, `addBorder = TRUE` adds zero-valued layers at the boundaries of the binning grid in the X and Z directions. These layers facilitate the closure of the isosurface during the Marching Cubes reconstruction.

```{r create-mesh}
object <- cloud[component == "object", ]
mesh <- createMesh(cloud = object, rBin = c(8, 8, 4), lowTol = 1e-2, trunc = 3, 
                   cad = cad, addBorder = TRUE)
```

The resulting triangular mesh represents the reconstructed surface of the object.

```{r show-reconstructed-mesh, rgl=TRUE}
# Represent the reconstructed object surface
# Use `open3d()` or `new3d()` to open a new device
shade3d(mesh, col = "lightgray")
```

```{r include=FALSE}
clear3d()
```

## Classification of object components

The reconstructed mesh can be further processed to identify relevant geometric components according to the characteristics of the object under study. In this example, the object contains a base and several reinforcements, which are identified according to the orientation and spatial distribution of the triangular faces.

Face normals and barycentres are first computed for each triangle.

```{r compute-face-geometry}
# Compute face normals and barycentres
normals <- vcgFaceNormals(mesh)
bary <- vcgBary(mesh)
```

### Identification of vertical faces

The orientation of each face relative to the X axis is characterized by the angle between its normal vector and this axis. Vertical faces are identified using an angular tolerance.

```{r classify-vertical-faces}
# Compute angles relative to the X axis
angleX <- angleAxis(v = normals, dim = 1, negative = TRUE)

# Angular tolerance
tolAngleX <- 60

# Identify vertical faces
isVertical <- angleX < tolAngleX | angleX > (180 - tolAngleX)
```

The resulting classification is shown below.

```{r show-vertical-faces, rgl=TRUE}
# Represent vertical faces
fshade3d(mesh, as.factor(isVertical), lit = FALSE)
```

### Identification of horizontal faces

The orientation of each face relative to the Z axis is characterized by the angle between its normal vector and this axis. Horizontal faces are identified using a corresponding angular tolerance.

```{r classify-horizontal-faces}
# Compute angles relative to the Z axis
angleZ <- angleAxis(v = normals, dim = 3)

# Angular tolerance
tolAngleZ <- 40

# Identify horizontal faces
isHorizontal <- angleZ < tolAngleZ
```

The resulting classification is shown below.

```{r show-horizontal-faces, rgl=TRUE}
# Represent horizontal faces
fshade3d(mesh, as.factor(isHorizontal), lit = FALSE)
```

### Identification of the object base

The object base is identified by selecting horizontal faces whose barycentres lie below a specified Z threshold.

```{r classify-base-faces}
# Identify faces belonging to the object base
tolZ <- 5
isBase <- isHorizontal & bary[, 3] < tolZ
```

The faces satisfying both conditions are represented below.

```{r show-base-faces, rgl=TRUE}
# Represent object-base faces
fshade3d(mesh, as.factor(isBase), lit = FALSE)
```

```{r include=FALSE}
clear3d()
```

### Identification of reinforcements

The reinforcements are identified from the spatial distribution of the barycentres of approximately vertical faces along the X direction. The modes of this distribution define the X intervals associated with the individual reinforcements.

```{r identify-reinforcements}
# Identify the X ranges associated with the reinforcements
modesX <- findModes(x = bary[isVertical, 1], bw = 8, q1 = 0.97)

# Number of detected reinforcements
nReinforcements <- nrow(modesX)
nReinforcements
```
A separate mesh is then extracted for each reinforcement. Since the selected triangles may form disconnected components, `splitTrianglesInd()` is used to identify connected groups of triangles. The `filterMeshComponents()` function is then applied to remove small isolated components, retaining the relevant geometry of each reinforcement.

```{r create-reinforcement-meshes}
reinforcementMesh <- vector("list", nReinforcements)
reinforcementInd <- vector("list", nReinforcements)
for (i in seq_len(nReinforcements)) {
  # Select triangles within the X range of the current reinforcement
  ind <- bary[, 1] > modesX[i, "left"] & bary[, 1] < modesX[i, "right"] &
         (isVertical | bary[, 3] > tolZ)

  # Create the candidate reinforcement mesh
  candidateMesh <- mesh
  candidateMesh$it <- mesh$it[, ind]

  # Identify connected triangle components
  components <- splitTrianglesInd(mesh = candidateMesh)

  # Remove small disconnected components
  filtered <- filterMeshComponents(mesh = candidateMesh, comps = components,
                                   minSize = 1000)

  # Store triangle indices in the original mesh
  reinforcementInd[[i]] <- which(ind)[filtered$triIdx]

  # Store the filtered reinforcement mesh
  reinforcementMesh[[i]] <- filtered$mesh
}
```

### Extraction of the object base

The object base is extracted using the same connected-component filtering procedure.

```{r create-base-mesh}
# Create the candidate base mesh
baseMesh <- mesh
baseMesh$it <- mesh$it[, isBase]

# Identify connected triangle components
components <- splitTrianglesInd(mesh = baseMesh)

# Remove small disconnected components
filtered <- filterMeshComponents(mesh = baseMesh, comps = components, minSize = 1000)

# Store the filtered base mesh
baseMesh <- filtered$mesh

# Store triangle indices in the original mesh
baseInd <- which(isBase)[filtered$triIdx]
```

### Representation of the classified components

The resulting `baseMesh` object contains the object base, while `reinforcementMesh` contains the individual reinforcements. The vectors `baseInd` and `reinforcementInd` retain their corresponding triangle indices in the original reconstructed mesh.

The classified structural components are represented together below.

```{r show-object-components, rgl=TRUE}
# Represent the object base and reinforcements
# Use `open3d()` or `new3d()` to open a new device

# Represent the object base
shade3d(baseMesh, col = "lightgray")

# Represent the reinforcements
for (i in seq_len(nReinforcements)) {
  shade3d(reinforcementMesh[[i]], col = "darkgreen", lit = FALSE)
}
```

```{r include=FALSE}
clear3d()
```

## Classification of reinforcement surfaces

Each reinforcement is further classified into its main surfaces. The bulb orientation
is first determined with `sideBulb()`. Face normals and their angles relative to the
X and Z axes are then used to identify the flat face of the reinforcement and the upper
flat face of its bulb.

```{r classify-reinforcement-surfaces}
# Determine the bulb orientation of each reinforcement
# TRUE = left, FALSE = right
bulbSide <- sapply(reinforcementMesh, sideBulb)
bulbSide

# Angular tolerance for the flat faces
tolAngleZFlat <- 60

reinforcementFlatMesh <- vector("list", nReinforcements)
bulbUpperMesh <- vector("list", nReinforcements)
for (i in seq_len(nReinforcements)) {
  # Compute face normals and barycentres
  normals <- vcgFaceNormals(reinforcementMesh[[i]])
  bary <- vcgBary(reinforcementMesh[[i]])

  # Compute angles relative to the X and Z axes
  angleX <- angleAxis(v = normals, dim = 1, negative = TRUE)
  angleZ <- angleAxis(v = normals, dim = 3)

  # Identify the flat face according to the bulb orientation
  if (bulbSide[i]) {
    isFlatFace <- angleX >= (180 - tolAngleX)
  } else {
    isFlatFace <- angleX < tolAngleX
  }

  isFlatFace <- isFlatFace & angleZ >= tolAngleZFlat

  # Create the candidate flat-face mesh
  candidateFlatMesh <- reinforcementMesh[[i]]
  candidateFlatMesh$it <- candidateFlatMesh$it[, isFlatFace]

  # Identify connected components
  components <- splitTrianglesInd(mesh = candidateFlatMesh)

  # Remove small disconnected components
  filtered <- filterMeshComponents(mesh = candidateFlatMesh, comps = components, 
                                   minSize = 2000)

  # Store the flat face of the reinforcement
  reinforcementFlatMesh[[i]] <- filtered$mesh

  # Determine the upper Z region of the reinforcement
  zLimit <- quantile(bary[, 3], probs = 0.8)

  # Identify the upper flat face of the bulb
  isBulbUpperFace <- angleZ <= tolAngleZ & bary[, 3] >= (zLimit + 5)

  ## Create the upper flat face of the bulb mesh
  bulbUpperMesh[[i]] <- reinforcementMesh[[i]]
  bulbUpperMesh[[i]]$it <- bulbUpperMesh[[i]]$it[ , isBulbUpperFace]
}
```

The resulting meshes identify two relevant surfaces of each reinforcement: the flat
face of the reinforcement and the upper flat face of its bulb. The bulb orientation
determined by `sideBulb()` is used to select the appropriate flat face of each reinforcement.

The classified surfaces of all reinforcements are represented together below.

```{r show-reinforcement-surfaces, rgl=TRUE}
# Represent the classified reinforcement surfaces
# Use `open3d()` or `new3d()` to open a new device
for (i in seq_len(nReinforcements)) {
  shade3d(reinforcementFlatMesh[[i]], col = "darkgreen", lit = FALSE)
  shade3d(bulbUpperMesh[[i]], col = "red", lit = FALSE)
}
```

```{r include=FALSE}
clear3d()
```

## Extraction of the object base boundary

The boundary of the object base provides geometric information for subsequent dimensional analysis. In this section, the boundary is identified, cleaned, and ordered to facilitate the identification of its corners and the segmentation of the contour into its four sides.

### Extraction of the original boundary

The original boundary of `baseMesh` is identified using `getBoundarySegments()`. In this example, setting `returnMesh = FALSE` returns a matrix containing the vertex indices defining each boundary edge. These edges include those associated with the intersections between the base and the reinforcements.

```{r extract-base-boundary}
# Extract the original boundary vertex indices
iBoundary <- getBoundarySegments(mesh = baseMesh, returnMesh = FALSE, simplify = FALSE)
```

The original boundary is represented below by connecting the corresponding pairs of vertices in `baseMesh`.

```{r show-original-base-boundary, rgl=TRUE}
# Represent the original boundary
shade3d(baseMesh, col = "lightgray", alpha = 0.4)
shade3d(mesh3d(vertices = baseMesh$vb, segments = iBoundary), col = "red")
```

```{r include=FALSE}
clear3d()
```

### Cleaning and ordering the boundary

Before cleaning the boundary, the vertices shared by the base and each reinforcement are identified from their triangle indices in the reconstructed mesh. These intersection vertices are then provided to `cleanBoundarySegments()`, which removes segments associated with the reinforcements, eliminates internal noise and irregular connections, and reconnects the resulting gaps to obtain a clean, closed boundary.

```{r clean-base-boundary}
# Identify vertex indices shared by the base and each reinforcement
baseReinforcementIntersection <- vector("list", nReinforcements)
for (i in seq_len(nReinforcements)) {
  baseReinforcementIntersection[[i]] <- intersect(
    mesh$it[, baseInd], mesh$it[, reinforcementInd[[i]]])
}

# Clean and reconnect the boundary segments
iBoundary <- cleanBoundarySegments(iBorder = iBoundary, baseMesh = baseMesh,
                                   intersection = baseReinforcementIntersection)
```

The cleaned boundary is represented below.

```{r show-clean-base-boundary, rgl=TRUE}
# Represent the cleaned boundary
shade3d(baseMesh, col = "lightgray", alpha = 0.4)
shade3d(mesh3d(vertices = baseMesh$vb, segments = iBoundary), col = "red")
```

```{r include=FALSE}
clear3d()
```

The cleaned boundary edges are subsequently ordered using `sortSegments()`, which arranges them into a continuous sequence according to their connectivity. This ordering facilitates the identification of the base corners and the segmentation of the contour into its four sides.

```{r sort-base-boundary}
# Order the cleaned boundary edges
iBoundary <- sortSegments(edges = iBoundary)
```

### Identification of the base corners

In this example, the base is approximately rectangular and aligned with the X and Y axes. The four corners are identified using the `detectCorners()` function from dimControl.

This function computes the convex hull of the boundary vertices and defines four reference corners using the minimum and maximum X and Y coordinates. The closest convex hull vertex to each reference corner is then selected using Euclidean distance. The resulting corners are labelled A (bottom-left), B (bottom-right), C (top-right), and D (top-left).

```{r identify-base-corners}
# Extract the ordered boundary vertex indices
iBoundaryVertices <- iBoundary[1, ]

# Extract the XY coordinates of the boundary vertices
boundaryXY <- t(baseMesh$vb[1:2, iBoundaryVertices])

# Identify the four corners
iCorner <- detectCorners(boundaryXY)

# Extract the corresponding mesh vertex indices
iCornerVertices <- iBoundaryVertices[iCorner]
```

The vector `iCorner` contains the indices of the four identified corners in the ordered boundary sequence, whereas `iCornerVertices` contains their corresponding vertex indices in `baseMesh`.

The identified corners are represented below on the cleaned and ordered boundary.

```{r show-base-corners, rgl=TRUE}
# Represent the cleaned and ordered boundary with the identified corners
shade3d(baseMesh, col = "lightgray", alpha = 0.4)
shade3d(mesh3d(vertices = baseMesh$vb, segments = iBoundary), col = "red")
points3d(t(baseMesh$vb[1:3, iCornerVertices]), size = 8)
```

```{r include=FALSE}
clear3d()
```

### Segmentation of the base boundary

Once the corners have been identified, their indices in the ordered boundary sequence are used to divide the contour into its four sides. Each side consists of consecutive boundary edges between two adjacent corners, including the side connecting the last corner to the first.

```{r segment-base-boundary}
# Order the corner positions without modifying their labels
iCornerOrdered <- sort(iCorner)
n <- ncol(iBoundary)

# Identify the boundary edge indices for each side
sideLimits <- list(
  iCornerOrdered[1]:(iCornerOrdered[2] - 1),
  iCornerOrdered[2]:(iCornerOrdered[3] - 1),
  iCornerOrdered[3]:(iCornerOrdered[4] - 1),
  c(iCornerOrdered[4]:n, seq_len(iCornerOrdered[1] - 1))
)

# Extract the vertex indices defining the edges of each side
boundarySides <- lapply(sideLimits, function(ind) iBoundary[, ind])
```

The four sides of the base boundary are represented below in different colors.

```{r show-segmented-base-boundary, rgl=TRUE}
# Represent the four sides of the base boundary
shade3d(baseMesh, col = "lightgray", alpha = 0.4)

color <- c("red", "blue", "green", "orange")
for (i in seq_len(4)) {
  shade3d(mesh3d(vertices = baseMesh$vb, segments = boundarySides[[i]]),
                 col = color[i], lwd = 4)
}
```

## Dimensional inspection results

The reconstructed mesh can be used to evaluate the dimensions of the base against the reference CAD model. In this example, the nominal CAD dimensions are known, and all coordinates and lengths are expressed in millimetres (mm).

### Corner coordinate comparison

The coordinates of the four identified corners are compared with their nominal positions in the CAD model. The differences are calculated as the reconstructed mesh coordinates minus the CAD coordinates.

```{r corner-comparison}
# Extract mesh corner coordinates
coord <- baseMesh$vb[1:3, iCornerVertices]
rownames(coord) <- c("X", "Y", "Z")
colnames(coord) <- names(iCorner)

# Define nominal CAD coordinates
cornerCAD <- rbind(
  X = c(A = 0, B = 900, C = 900, D = 0),
  Y = c(A = 0, B = 0, C = 500, D = 500),
  Z = c(A = 0, B = 0, C = 0, D = 0)
)

# Compare CAD and mesh coordinates
cornerTable <- data.frame(
  Corner = colnames(coord),
  X_CAD = cornerCAD["X", ],
  Y_CAD = cornerCAD["Y", ],
  Z_CAD = cornerCAD["Z", ],
  X_Mesh = coord["X", ],
  Y_Mesh = coord["Y", ],
  Z_Mesh = coord["Z", ],
  DX = coord["X", ] - cornerCAD["X", ],
  DY = coord["Y", ] - cornerCAD["Y", ],
  DZ = coord["Z", ] - cornerCAD["Z", ],
  row.names = NULL
)

knitr::kable(cornerTable, digits = 3, caption = "Comparison of CAD and mesh corner coordinates (mm).")
```

The reconstructed base is approximately 3.708 mm above the CAD plane (Z = 0). This difference may be caused by the reconstruction process. The mesh has not been shifted or aligned with the CAD model.

```{r plot-corner-deviations, fig.width=8, fig.height=4}
# Corner deviations 
cornerDeviations <- cornerTable[, c("Corner", "DX", "DY", "DZ")]

# Plot corner coordinate deviations
ggplot2::ggplot(
  tidyr::pivot_longer(cornerDeviations, DX:DZ),
  ggplot2::aes(x = Corner, y = value, fill = name)
) +
  ggplot2::geom_col(position = "dodge") +
  ggplot2::geom_hline(yintercept = 0) +
  ggplot2::scale_fill_manual(values = c("steelblue", "orange", "seagreen")) +
  ggplot2::labs(
    title = "Corner coordinate deviations",
    x = "Corner",
    y = "Deviation (mm)",
    fill = "Coordinate"
  ) +
  ggplot2::theme_minimal()
```

### Base dimensions

The lengths of the four sides are calculated as the Euclidean distances between adjacent corners using `euclideanDistance()`. These lengths are compared with the known nominal CAD dimensions.

```{r base-dimensions}
# Calculate the side lengths
AB <- euclideanDistance(coord[, "B"] - coord[, "A"])
BC <- euclideanDistance(coord[, "C"] - coord[, "B"])
CD <- euclideanDistance(coord[, "D"] - coord[, "C"])
DA <- euclideanDistance(coord[, "A"] - coord[, "D"])

# Compare mesh and CAD dimensions
baseDimensions <- data.frame(
  Side = c("AB", "BC", "CD", "DA"),
  Mesh = c(AB, BC, CD, DA),
  CAD = c(900, 500, 900, 500),
  Difference = c(AB, BC, CD, DA) - c(900, 500, 900, 500)
)

knitr::kable(baseDimensions, digits = 3,
  col.names = c("Side", "Mesh (mm)", "CAD (mm)", "Difference (mm)"),
  caption = "Comparison of mesh and nominal CAD dimensions."
)
```

Positive differences indicate dimensions larger than the nominal CAD values, whereas negative differences indicate smaller dimensions.


```{r plot-dimensional-deviations, fig.width=8, fig.height=4}
# Plot side length deviations
ggplot2::ggplot(baseDimensions,
                ggplot2::aes(x = Side, y = Difference)) +
  ggplot2::geom_col(fill = "steelblue", width = 0.6) +
  ggplot2::geom_text(
    ggplot2::aes(label = sprintf("%+.3f", Difference)),
    vjust = -0.5
  ) +
  ggplot2::labs(
    title = "Side length deviations from the CAD model",
    x = "Side",
    y = "Length deviation (mm)"
  ) +
  ggplot2::theme_minimal()
```


## Further help

See the function [reference pages](https://modes-cemi.github.io/dimControl/reference/) for the full list of arguments and additional examples.
